Analytic Doppler Imaging

### 2019-08-22 tex/ms.tex
more text [build paper]

\begin{figure}[t!]
    \begin{centering} 
    \includegraphics[width=0.85\linewidth]{figures/spot_setup.pdf}
    \oscaption{spot_setup}{%
        The basic setup for the \spot problem. We compute the 
        $l_\mathrm{max} = 15$ spherical harmonic expansion of 
        a map where the word ``SPOT'' has been placed across the northern 
        hemisphere (top left). Each point on the surface is
        given the same spectrum, composed of a sum of synthetic 
        Gaussian absorption lines (blue curve in the bottom panel). 
        The specific intensity
        at each point on the surface is equal to this spectrum weighted
        by the intensity of the map, which is close to zero within the
        spot (point labeled \texttt{B}) and close ot unity 
        outside the spot (point labeled \texttt{A}).
        %
        To generate the synthetic dataset (orange curves in the bottom
        panel), the star is inclined at
        $40^\circ$ with respect to the line of sight and the disk-integrated
        spectrum is computed at 16 equally-spaced phases between
        $-180^\circ$ and $180^\circ$ (middle panel).
        %
        \label{fig:spot_setup}
    }
    \end{centering}
\end{figure}

@@

Given $\bvec{y_1}$ and $\bvec{s}$, we compute the data vector $\bvec{f}$
at $M = 16$ equally spaced rotational phases; these phases are shown in orthographic
projection (as an observer would see the star) in the center
panel of Figure~\ref{fig:spot_setup}. The synthetic spectra for each phase
are shown as the orange curves in the bottom panel of the figure (note the
different scaling, as all spectral lines become shallower due to the
broadening). We assume a wavelength-independent,
quadratic limb darkening law with $u_1 = 0.5$ and $u_2 = 0.25$. Finally,
we add random Gaussian noise with amplitude $\sigma_\bvec{f} = 10^{-4}$ to
simulate a very high signal-to-noise ratio (SNR) spectrum. The mean
standard deviation in each wavelength bin across all 16 epochs in the
noiseless data vector is about $10^{-2}$, so our choice of $\sigma_\bvec{f}$
corresponds to SNR $\sim$ 100 in each of the 201 wavelength bins of $\bvec{f}$.
%
%For reference, \citet{Vogt1987} report SNRs per 60 m\AA\, pixel, while our
%resolution elements are approximately $\lambda_r\Delta\lnlam/K \approx 100$ m\AA\,
%across, so our SNR in the same units as \citet{Vogt1987} is about 80.
%
% TODO: Perhaps Vogt computes the "signal" as the line depth, rather than
% as the amplitude of the perturbations? Check this.

@@

\subsection{Known baseline and spectrum}
In our first experiment, we attempt to infer the stellar map...

@@

@@

\begin{figure}[p!]
    \begin{centering}
    \includegraphics[width=0.49\linewidth]{figures/spot_y1bs_timeseries.pdf}
    \includegraphics[width=0.49\linewidth]{figures/spot_y1bs_rect.pdf}
    \\[0.5em]
    \includegraphics[width=0.405\linewidth]{figures/spot_y1bs_baseline.pdf}
    \includegraphics[width=0.575\linewidth]{figures/spot_y1bs_coeffs.pdf}
    \\[0.5em]
    \includegraphics[width=0.98\linewidth]{figures/spot_y1bs_spectrum.pdf}
    \oscaption{spot_y1bs}{%
        Similar to Figure~\ref{fig:spot_y1}, but assuming we have no
        information about either the spectrum, the baseline, or the map
        coefficients. The bottommost panel now shows the true rest frame
        spectrum (blue line), the initial guess based on a deconvolution
        (dotted orange line), and the inferred spectrum (orange line, with
        shaded $1\sigma$ uncertainty). The grey regions indicate wavelength
        regions outside the observation window, but whose features still
        contribute to the observed spectrum due to rotational broadening.
        \label{fig:spot_y1bs}
    }
    \end{centering}
\end{figure}

>>>>

\begin{figure}[t!]
    \begin{centering} 
    \includegraphics[width=0.85\linewidth]{figures/spot_setup.pdf}
    \oscaption{spot_setup}{%
        The basic setup for the \spot problem. We compute the 
        $l_\mathrm{max} = 15$ spherical harmonic expansion of 
        a map where the word ``SPOT'' has been placed across the northern 
        hemisphere (top left). Each point on the surface is
        given the same spectrum, composed of a sum of synthetic 
        Gaussian absorption lines (blue curve in the bottom panel). 
        The specific intensity
        at each point on the surface is equal to this spectrum weighted
        by the intensity of the map, which is close to zero within the
        spot (point labeled \texttt{B}) and close ot unity 
        outside the spot (point labeled \texttt{A}).
        %
        To generate the synthetic dataset (orange curves in the bottom
        panel), the star is inclined at
        $40^\circ$ with respect to the line of sight and the disk-integrated
        spectrum is computed at 16 equally-spaced phases between
        $-180^\circ$ and $180^\circ$ (middle panel).
        %
        The grey regions indicate wavelength
        regions outside the observation window, but whose features still
        contribute to the observed spectrum due to rotational broadening.
        \label{fig:spot_setup}
    }
    \end{centering}
\end{figure}

@@

Given $\bvec{y_1}$ and $\bvec{s}$, we use Equation~(\ref{eq:deconv:F}) to
compute the data vector $\bvec{f}$
at $M = 16$ equally spaced rotational phases; these phases are shown in orthographic
projection (as an observer would see the star) in the center
panel of Figure~\ref{fig:spot_setup}. The synthetic spectra for each phase
are shown as the orange curves in the bottom panel of the figure (note the
different scaling, as all spectral lines become shallower due to the
broadening). We assume a wavelength-independent,
quadratic limb darkening law with $u_1 = 0.5$ and $u_2 = 0.25$. Finally,
we divide the flux by the baseline at each epoch (Equation~\ref{eq:norm:b})
and add random Gaussian noise with amplitude $\sigma_\bvec{f} = 10^{-4}$ to
simulate a very high signal-to-noise ratio (SNR) spectrum. The mean
standard deviation in each wavelength bin across all 16 epochs in the
noiseless data vector is about $10^{-2}$, so our choice of $\sigma_\bvec{f}$
corresponds to SNR $\sim$ 100 in each of the 201 wavelength bins of $\bvec{f}$.
%
%For reference, \citet{Vogt1987} report SNRs per 60 m\AA\, pixel, while our
%resolution elements are approximately $\lambda_r\Delta\lnlam/K \approx 100$ m\AA\,
%across, so our SNR in the same units as \citet{Vogt1987} is about 80.
%
% TODO: Perhaps Vogt computes the "signal" as the line depth, rather than
% as the amplitude of the perturbations? Check this.

@@

In the following sections we discuss various approaches to 
solving the Doppler imaging problem for this dataset.
Unless otherwise noted, in all experiments we adopt a Gaussian prior
for the spherical harmonic coefficients $\bvec{y_1}$ with mean
$\boldsymbol{\mu}_\bvec{y_1} = \bvec{0}$ and covariance
$\boldsymbol{\Lambda}_\bvec{y_1} = (10^{-2})^2\bvec{I}$,
corresponding to a flat power spectrum prior.
We place a Gaussian process prior on the spectrum $\bvec{s}$ with mean
$\boldsymbol{\mu}_\bvec{s} = \bvec{1}$ and a covariance matrix
$\boldsymbol{\Lambda}_\bvec{s}$ computed from
 a Mat\'ern-3/2 kernel with amplitude $\sigma_\bvec{s} = 0.3$
and lengthscale $\rho_\bvec{s} = 3\times 10^{-5}$.

\subsection{Known baseline and spectrum}
%
In our first experiment, we attempt to infer the stellar map assuming
we have perfect knowledge of the rest frame spectrum $\bvec{s}$ and
the baseline $\bvec{b}$ (Equation~\ref{eq:norm:b}). 
This is analogous to the case where a good physical model exists for
the stellar spectrum and simultaneous photometry is available to
constrain the baseline variations.
As we showed in
\S\ref{sec:solve_y}, the posterior distribution for the map coefficients 
$\bvec{y_1}$ is analytic in this case, with mean given by Equation~(\ref{eq:inverse:y1hat})
and covariance given by Equation~(\ref{eq:inverse:y1cov}). Note that
we must first scale the data vector $\bvec{f}$ by $\bvec{b}$ (which in this
case is known).

@@

The results are shown in Figure~\ref{fig:spot_y1}. The left hand panel
shows the inferred map alongside the data
(black points) and best fit model (orange curves) at each of the 16 epochs.
The panels on the right show the true map (top), the inferred map (center)
and the posterior pointwise uncertainty (bottom) on a rectangular
latitude-longitude grid. The uncertainty on the vector of pixels $\bvec{p}$ where
we evaluate the map is equal to the square root of the diagonal of
the posterior covariance matrix for the pixel values, computed from
%
\begin{align}
    \label{eq:spotstar:uncert}
    \boldsymbol{\Sigma}_\bvec{p} = \bvec{P} \boldsymbol{\Sigma}_\bvec{\hat{y}_1}^{-1} \bvec{P}^\top
\end{align}
%
where $\bvec{P}$ is the change-of-basis matrix from spherical harmonic 
coefficients to pixels on the grid.

The inferred map is nearly indistinguishable from the true map, save for a
few small differences, particularly below the stellar equator. This is 
captured by the uncertainty, which increases
south of $+30^\circ$ latitude, reaching a maximum at $-30^\circ$, below
which the uncertainty is constant at about 10\%. This is due to the fact
that regions below $-40^\circ$ latitude are never in view from the
observer's perspective, while regions between about $-30^\circ$ and
$-40^\circ$ are so strongly limb-darkened that the data contains little
information about them. The particular value of the uncertainty in this
region---10\%---is the uncertainty associated with the prior, which can
be computed by replacing $\boldsymbol{\Sigma}_\bvec{\hat{y}_1}$ with 
$\boldsymbol{\Lambda}_\bvec{y_1}$ in Equation~(\ref{eq:spotstar:uncert}).

\subsection{Known spectrum, unknown baseline}
%
We next investigate the case where the spectrum is known exactly, but
the baseline is not.

@@

\begin{figure}[p!]
    \begin{centering}
    \includegraphics[width=0.49\linewidth]{figures/spot_y1bs_timeseries.pdf}
    \includegraphics[width=0.49\linewidth]{figures/spot_y1bs_rect.pdf}
    \\[0.5em]
    \includegraphics[width=0.405\linewidth]{figures/spot_y1bs_baseline.pdf}
    \includegraphics[width=0.575\linewidth]{figures/spot_y1bs_coeffs.pdf}
    \\[0.5em]
    \includegraphics[width=0.98\linewidth]{figures/spot_y1bs_spectrum.pdf}
    \oscaption{spot_y1bs}{%
        Similar to Figure~\ref{fig:spot_y1}, but assuming we have no
        information about either the spectrum, the baseline, or the map
        coefficients. The bottommost panel now shows the true rest frame
        spectrum (blue line), the initial guess based on a deconvolution
        (dotted orange line), and the inferred spectrum (orange line, with
        shaded $1\sigma$ uncertainty).
        \label{fig:spot_y1bs}
    }
    \end{centering}
\end{figure}
